### 2024-06-07 products/manuscript/Manuscript.qmd
first update

---
title: "Predicting and Preventing Employee Turnover through HR Analytics"
subtitle: ""
author: Guillermo Gallardo
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/apa.csl
---

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

{{\< pagebreak \>}

# Introduction

## General Background Information

The primary objective of this project is to identify the key factors contributing to employee turnover and to develop a model that can predict potential departures within a company. By anticipating these departures, we can create strategies for employee retention and enhance the overall employee experience.

I find this dataset interesting, as it aligns with my previous experience in HR Analytics. Although I've worked in this field before, I never had the opportunity to work on project where we could predict attrition. This project allows me to deep dive into the data I am interested about, while also enhancing my skills in modeling using R.

## Datasets

**IBM HR Analytics Employee Attrition**

The dataset for this project was sourced from [Kaggle.com](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset) and was created by IBM Data Scientists as a fictional representation of HR data. It consists of 1,470 rows and 35 variables that cover demographic details, job characteristics, compensation, performance metrics, and employee satisfaction data. This data serves as the foundation for analyzing factors contributing to employee turnover and developing a predictive model.

**HR Analytics Case Study**

This data set was also sources from [Kaggle.com](https://www.kaggle.com/datasets/vjchoudhary7/hr-analytics-case-study) and the collaborators are Vijay Choudhary and Aman Kumar. This data set consists of 5 data files and 1 data dictionary file. Each of the data files, contains 4,410 rows and the main file contains 24 variables similar to the IBM file where it shows demographic information, job characteristics, compensation, survey data, etc. This file contains additional information that could help us

## Concerns

The data

*DELETE ----- Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*

## Questions/Hypotheses to be addressed

*State the research questions you plan to answer with this analysis.*

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2024-06-07 products/manuscript/Manuscript.qmd
Second update with code.

@@
 library(knitr)
+
+hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE)
 ```
 
-{{\< pagebreak \>}
-
-# Introduction
-
-## General Background Information
+# Project Overview
 
@@
 
-## Datasets
+# Project Question
 
-**IBM HR Analytics Employee Attrition**
+We have two main questions that we want to answer during this project.
+
+-   What are the primary factors that contribute to employee attrition?
+
+-   What strategies can be implement to enhance employee retention and improve the overall employee experience?
+
+# Dataset
+
+For this project, I was looking for a dataset with employee data that included a wide range of variables to explore potential relationships with attrition.
+Below, I will provide an overview of the dataset's origin, author, the types of variables it includes, and other relevant details.
+
+## **IBM HR Analytics Employee Attrition**
 
@@
 
-**HR Analytics Case Study**
+## **Variables**
 
-This data set was also sources from [Kaggle.com](https://www.kaggle.com/datasets/vjchoudhary7/hr-analytics-case-study) and the collaborators are Vijay Choudhary and Aman Kumar.
-This data set consists of 5 data files and 1 data dictionary file.
-Each of the data files, contains 4,410 rows and the main file contains 24 variables similar to the IBM file where it shows demographic information, job characteristics, compensation, survey data, etc.
-This file contains additional information that could help us
+This file includes 35 variables, detailed below, which includes a combination of character and integer types, with some of the character variables being categorical.
 
-## Concerns
+```{r Variables}
 
-The data
+str(hr_data)
+```
 
-*DELETE ----- Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section.*
+## **Gender**
 
-## Questions/Hypotheses to be addressed
+Out of the 1,470 observation in the dataset, 588 are female and 882 are male. This provides a quick snapshot of the gender distribution within the data.
 
-*State the research questions you plan to answer with this analysis.*
+```{r Gender}
+table(hr_data$Gender)
+```
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
+## **Education**
 
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+During data cleaning, we'll use the data dictionary to decode each field.
+For instance, the education field is labeled from 1 to 5, representing different educational levels: 1 = "Below College," 2 = "College," 3 = "Bachelor's," 4 = "Master's," and 5 = "Doctorate." Notably, about 40% of the dataset holds a Bachelor's degree.
 
-{{< pagebreak >}}
+```{r Education}
+table(hr_data$Education)
+```
 
@@
 
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
-
-## Data aquisition
-
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
-
-## Data import and cleaning
-
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+## Exploratory/Descriptive analysis
 
@@
 *Explain anything related to your statistical analyses.*
-
-{{< pagebreak >}}
-
-# Results
-
-## Exploratory/Descriptive analysis
-
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
-
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
-
-## Basic statistical analysis
-
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
-## Full analysis
-
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
-
-{{< pagebreak >}}
-
-# Discussion
-
-## Summary and Interpretation
-
-*Summarize what you did, what you found and what it means.*
-
-## Strengths and Limitations
-
-*Discuss what you perceive as strengths and limitations of your analysis.*
-
-## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
-
-{{< pagebreak >}}
-
-# References

### 2024-06-07 products/manuscript/Manuscript.qmd
.

@@
 
-hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE)
+hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE) #not synced with OneDrive but my computer added that folder when I entered my email when I bought the computer
 ```

### 2024-06-07 products/manuscript/Manuscript.qmd
Updating analysis methods

@@
 
-# Methods
+# Analysis Methods
 
-*Describe your methods. That should describe the data, the cleaning processes, and the analysis approaches. You might want to provide a shorter description here and all the details in the supplement.*
+## Data Cleaning Process
 
-## Exploratory/Descriptive analysis
+In the data cleaning process, I will use the data dictionary to enhance data interpretation.
+For example, I will address issues like the Education field being labeled from 1 to 5 instead of using descriptive education level labels.
+I will also examine variables to detect any inaccuracies.
+An example of that could be to exclude the compensation fields (hourly, daily, and monthly rates) and use the Monthly Income field to ensure consistency and clarity.
 
-## Statistical analysis
+## Descriptive Analysis
 
-*Explain anything related to your statistical analyses.*
+I will begin my project descriptive analysis, with the goal of summarizing the data.
+This first step will provide insights into any outliers or patterns within the dataset.
+This information will also help us with creating visualization that will highlight the patterns in the data so we can focus on specific areas.
+
+## Correlation Analysis
+
+As we explore various analysis methods in our class, my focus is on finding potential correlation between variables and employee attrition.
+I am interested in using the demographic details, job attributes and survey data to uncover pattern (if any) that highlight this issue .
+
+## Predictive Modeling
+
+My main goal is to predict when an employee might leave the company.
+I am interested in using predictive modeling techniques (also part of my other summer class) to see if we can predict attrition.
+Once I've identified potential departures, we can use this findings to create strategies to prevent or reduce attrition.

### 2024-06-14 code/eda-code/HReda.qmd
Project part 2 - first update

---
title: "HR Data Exploratory Analysis Script"
date: "Summer 2024"
output: html_document
---

This Quarto file loads the cleaned data and does some exploring.

I'm only showing it the way where the code is included in the file. 
As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.

But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)

Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.

Also note that while here I split cleaning and exploring, this is iterative. You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it. In general, as you explore, you'll find things that need cleaning. As you clean, you can explore more. Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file. Or split things in any other logical way.

As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data. Depending on the total number of variables in your dataset, explore all or some of the others. Figures produced here might be histograms or density plots, correlation plots, etc. Tables might summarize your data.

Start by exploring one variable at a time. Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in. If your dataset is small, you can do that for all variables.

Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.

# Setup

```{r}
#load needed packages. make sure they are installed.
library(here) #for data loading/saving
library(dplyr)
library(skimr)
library(ggplot2)
```

Load the data.

```{r}
#Path to data. Note the use of the here() package and not absolute paths
data_location = here::here("data","processed-data","HRprocesseddata.rds")
#load data
myhrdata = readRDS(data_location)
```

# HR Data Exploration Through Tables

```{r}
col_names = names(myhrdata)
num_columns <- 3

num_rows <- ceiling(length(col_names) / num_columns)

summary_matrix <- matrix(col_names, nrow = num_rows, byrow = TRUE)

summary_df <- as.data.frame(summary_matrix)

# Print the summary data frame
print(summary_df)
#Save to file
summarytable_file = here("results", "tables", "HRvariablestable.rds")
saveRDS(summary_df, file = summarytable_file)

```

```{r}
summary_df = skimr::skim(myhrdata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "HRsummarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

Gender Table

```{r}
#summary_df = skimr::skim(myhrdata)
#print(summary_df)

summary_df = myhrdata %>%
  count(Gender) %>%
  mutate(`Total Count` = n,
         `% of Total` = (n / sum(n)) * 100) %>%
  select(Gender, `Total Count`, `% of Total`)

print(summary_df)
# save to file
summarytable_file = here("results", "tables", "HRgendertable.rds")
saveRDS(summary_df, file = summarytable_file)
```

Gender and Tenure Boxplot

```{r}
p1 = myhrdata %>%
  ggplot(aes(x = YearsAtCompany)) +
  geom_histogram(binwidth = 1, boundary = 0.5, fill = "gray", color = "black") +
  labs(x = "Tenure", y = "Count") + 
  theme_minimal() +
  theme(panel.background = element_rect(fill = "white"))

print(p1)

figure_file = here("results", "figures","tenure-distribution.png")
ggsave(filename = figure_file, plot=p1)
```

# Data exploration through tables

Showing a bit of code to produce and save a summary table.

```{r}
summary_df = skimr::skim(mydata)
print(summary_df)
# save to file
summarytable_file = here("results", "tables", "summarytable.rds")
saveRDS(summary_df, file = summarytable_file)
```

We are saving the results to the `results` folder. Depending on how many tables/figures you have, it might make sense to have separate folders for each. And/or you could have separate folders for exploratory tables/figures and for final tables/figures. Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.

# Data exploration through figures

Histogram plots for the continuous outcomes.

Height first.

```{r}
p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
plot(p1)
figure_file = here("results", "figures","height-distribution.png")
ggsave(filename = figure_file, plot=p1) 
```

Now weights.

```{r}
p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
plot(p2)
figure_file = here("results", "figures","weight-distribution.png")
ggsave(filename = figure_file, plot=p2) 
```

Now height as function of weight.

```{r}
p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
plot(p3)
figure_file = here("results", "figures", "height-weight.png")
ggsave(filename = figure_file, plot=p3) 
```

Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.

```{r}
p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
plot(p4)
figure_file = here("results", "figures", "height-weight-stratified.png")
ggsave(filename = figure_file, plot=p4) 
```

# Notes

For your own explorations, tables and figures can be "quick and dirty". As long as you can see what's going on, there is no need to polish them. That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.). Those should look as nice, polished and easy to understand as possible.

### 2024-06-14 code/processing-code/HRprocessingfile.qmd
Project part 2 - first update

---
title: "Cleaning HR Data for Analysis"
author: "Guillermo Gallardo"
date: "Summer 2024"
output: html_document
---

# Processing script

This contains the same code and comments/information as `processingcode.R`.

This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.

See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.

# Setup

Load needed packages. make sure they are installed.

```{r}
library(readxl) #for loading Excel files
library(dplyr) #for data processing/cleaning
library(tidyr) #for data processing/cleaning
library(skimr) #for nice visualization of data 
library(here) #to set paths
```

# Data loading

Note that for functions that come from specific packages (instead of base R), I often specify both package and function like so:
package::function() that's not required one could just call the function specifying the package makes it clearer where the function "lives",
but it adds typing. You can do it either way.

```{r}
# path to data
# note the use of the here() package and not absolute paths
hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx")
rawdata = readxl::read_excel(hr_data)
```

# Check data

First we can look at the codebook

```{r}
codebook = readxl::read_excel(hr_data, sheet ="Codebook")
print(codebook)
```

Several ways of looking at the data

```{r}
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

**NEED TO UPDATE THE EDUCATION FIELD**

# Cleaning HR Data

Removing three columns from dataset. This columns provide incorrect salary information. After reviewing the data, I decided to keep only the *MonthlyIncome* field to have consistency in the analysis.

```{r removing columns with incorrect data}
d1 = rawdata %>% 
  select(-DailyRate, -HourlyRate, -MonthlyRate)

head(d1)
```

The *Education* field was a numeric field that only showed 1-5 so I added labels to this so we can better understand our employees.
```{r adding label to Education}

d2 = d1 %>% 
    mutate(Education = case_when(
    Education == 1 ~ "Below College",
    Education == 2 ~ "College",
    Education == 3 ~ "Bachelor's",
    Education == 4 ~ "Master's",
    Education == 5 ~ "Doctorate",
    TRUE ~ as.character(Education)
  ))

head(d2)

```

```{r}
HRprocesseddata = d2
```

# Save data

Finally, we save the clean data as RDS file. I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
This preserves coding like factors, characters, numeric, etc.  If you save as CSV, that information would get lost.
However, CSV is better for sharing with others since it's plain text. If you do CSV, you might want to write down somewhere what each variable is.

See here for some suggestions on how to store your processed data:
http://www.sthda.com/english/wiki/saving-data-into-r-data-format-rds-and-rdata

```{r}
save_data_location = here::here("data","processed-data","HRprocesseddata.rds")
saveRDS(HRprocesseddata, file = save_data_location)
```

# Notes

Removing anyone who had "faulty" or missing data is one approach. It's often not the best. based on your question and your analysis approach, you might want to do cleaning differently (e.g. keep individuals with some missing information).

# Cleaning DELETE WHEN DONE

By inspecting the data as done above, we find some problems that need addressing:

First, there is an entry for height which says "sixty" instead of a number. 
Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
Since we don't know how to fix this, we might decide to remove the person. This "sixty" entry also turned all Height entries into characters instead of numeric. That conversion to character also means that our summary function isn't very meaningful. So let's fix that first.

```{r}
d1 <- rawdata %>% dplyr::filter( Height != "sixty" ) %>% 
                  dplyr::mutate(Height = as.numeric(Height))
skimr::skim(d1)
hist(d1$Height)
```

Now we see that there is one person with a height of 6. That could be a typo, or someone mistakenly entered their height in feet. Since we unfortunately don't know, we might need to remove this person, which we'll do here.

```{r}
d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height=="6",round(6*30.48,0)) )
skimr::skim(d2)
```

Height values seem ok now.

Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
To be able to analyze the data, we'll remove those individuals as well.

```{r}
d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
skimr::skim(d3)
```

Now checking the `Gender` variable. Gender should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.

```{r}
d3$Gender <- as.factor(d3$Gender)  
skimr::skim(d3)
```

Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Gender, I'm also using droplevels() to get rid of it.

```{r}
d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
skimr::skim(d4)
```

All done, data is clean now.

Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

### 2024-06-14 products/manuscript/Manuscript.qmd
Project part 2 - first update

@@
 
-hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE) #not synced with OneDrive but my computer added that folder when I entered my email when I bought the computer
+#hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE) #not synced with OneDrive but my computer added that folder when I entered my email when I bought the computer
 ```
@@
 
-## **Variables**
+# Data Cleaning Process
 
-This file includes 35 variables, detailed below, which includes a combination of character and integer types, with some of the character variables being categorical.
+Data cleaning is an important step in data analysis process and focuses on detecting and correcting errors and inconsistencies in the dataset.
+The goals is to ensure that the data is accurate, complete and reliable.
+Overall, this dataset didn't have any anomalies and there were only a couple of steps I took to prepare the data.
+The steps we took are the following:
 
-```{r Variables}
+**Loading Data for Cleaning Process**
 
-str(hr_data)
+In order to start the cleaning process, we need make to sure we load the data.
+In the code below, we use the *library(here)* function to set the file path dynamically, so our script can find and access the data files regardless of the current working directory.
+
+```{r, message=FALSE}
+#| echo: false
+library(here)
+library(dplyr)
 ```
 
-## **Gender**
-
-Out of the 1,470 observation in the dataset, 588 are female and 882 are male. This provides a quick snapshot of the gender distribution within the data.
-
-```{r Gender}
-table(hr_data$Gender)
+```{r, results='hide'}
+hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx")
+rawdata = readxl::read_excel(hr_data)
 ```
 
-## **Education**
+**Removing Column**
+
+In the next step of my data cleaning process, I removed three columns from the dataset.
+The columns we removed are *DailyRate*, *HourlyRate* and *MonthlyRate.* These columns were removed because the numbers they contained did not align logically or meaningfully.
+Upon further investigation, it became evident that there was no clear resolution to these discrepancies in Kaggle, and many suggested focusing instead on the Monthly Income field to maintain data consistency and reliability.
+
+```{r, results='hide'}
+d1 = rawdata %>% 
+  select(-DailyRate, -HourlyRate, -MonthlyRate)
+
+head(d1)
+```
+
+**Adding Labels**
+
+Next, we addressed the *Education* field as the third step in our process.
+Initially numeric, it contained values ranging from 1 to 5.
+Using our data dictionary, we updated these numeric values to their corresponding labels.
+This adjustment enhances our understanding of employee education levels in the dataset and ensures consistency in the data.
+
+```{r, results='hide'}
+d2 = d1 %>% 
+    mutate(Education = case_when(
+    Education == 1 ~ "Below College",
+    Education == 2 ~ "College",
+    Education == 3 ~ "Bachelor's",
+    Education == 4 ~ "Master's",
+    Education == 5 ~ "Doctorate",
+    TRUE ~ as.character(Education)
+  ))
+
+head(d2)
+```
+
+**Saving File**
+
+The final step in this process involved saving the file as an RDS file. This ensures that we can use it for our exploratory data analysis and maintain an organized project that is reproducible.
+
+```{r, results='hide'}
+HRprocesseddata = d2
+
+save_data_location = here::here("data","processed-data","HRprocesseddata.rds")
+saveRDS(HRprocesseddata, file = save_data_location)
+```
+
+# Analysis Methods
+
+## Exploratory Data Analysis
+
+I am starting my exploratory data analysis project with the objective of summarizing the dataset.
+I will begin by providing a concise overview of the variables included.
+Additionally, I will generate plots and tables to identify any outliers or patterns within the dataset.
+This analysis will help us understand the data available to us and guide our strategy to predict employee attrition within the organization.
+
+### Data Variables
+
+This dataset contains 35 variables, consisting of both character and integer types, with several character variables categorized as categorical.
+It provides valuable insights into employees, including demographic information such as age, gender, and marital status.
+Additionally, it offers details about their employment, including monthly income, department, and tenure, which are crucial for understanding their professional profiles and organizational dynamics.
+
+```{r}
+#| label: col-hrprocessingdata
+#| tbl-cap: "HR Data Variables"
+#| echo: FALSE
+variables=readRDS("../../results/tables/HRvariablestable.rds")
+knitr::kable(variables)
+```
+
+### **Gender**
+
+The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%.
+
+**Distribution**
+
+```{r}
+#| label: col-hrgender
+#| tbl-cap: "Employee Gender"
+#| echo: FALSE
+gender_table=readRDS("../../results/tables/HRgendertable.rds")
+knitr::kable(gender_table)
+```
+
+### **Employee Tenure (in years)**
+
+I examined the distribution of employee tenure and identified two distinct peaks: one at 5 years and another at 1 year.
+This provide quick insights into the typical duration of employees within the organization, and help us identify patterns that may influence employees to resign.
+
+```{r}
+#| label: fig-tenure
+#| fig-cap: "Employee Tenure"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","tenure-distribution.png"))
+```
+
+**MAYBE ADD LABEL FOR TENURE?**
+
+### **Education** MOVE TO EDA
 
@@
 
-```{r Education}
-table(hr_data$Education)
+```{r }
+
 ```
-
-# Analysis Methods
-
-## Data Cleaning Process
-
-In the data cleaning process, I will use the data dictionary to enhance data interpretation.
-For example, I will address issues like the Education field being labeled from 1 to 5 instead of using descriptive education level labels.
-I will also examine variables to detect any inaccuracies.
-An example of that could be to exclude the compensation fields (hourly, daily, and monthly rates) and use the Monthly Income field to ensure consistency and clarity.
-
-## Descriptive Analysis
-
-I will begin my project descriptive analysis, with the goal of summarizing the data.
-This first step will provide insights into any outliers or patterns within the dataset.
-This information will also help us with creating visualization that will highlight the patterns in the data so we can focus on specific areas.
 

### 2024-06-15 code/eda-code/HReda.qmd
Project part 2 - second update

@@
 ---
-
-This Quarto file loads the cleaned data and does some exploring.
-
-I'm only showing it the way where the code is included in the file. 
-As described in the `processing_code` materials, I currently prefer the approach of having R code in a separate file and pulling it in.
-
-But I already had this written and haven't yet re-done it that way. Feel free to redo and send a pull request on GitHub :)
-
-Again, it is largely a matter of preference and what makes the most sense to decide if one wants to have code inside Quarto files, or as separate R files.
-And sometimes, an R script with enough comments is good enough and one doesn't need a Quarto file.
-
-Also note that while here I split cleaning and exploring, this is iterative.
-You saw that as part of the processing, we already had to explore the data somewhat to understand how to clean it.
-In general, as you explore, you'll find things that need cleaning.
-As you clean, you can explore more.
-Therefore, at times it might make more sense to combine the cleaning and exploring code parts into a single R or Quarto file.
-Or split things in any other logical way.
-
-As part of the exploratory analysis, you should produce plots or tables or other summary quantities for the most interesting/important quantities in your data.
-Depending on the total number of variables in your dataset, explore all or some of the others.
-Figures produced here might be histograms or density plots, correlation plots, etc.
-Tables might summarize your data.
-
-Start by exploring one variable at a time.
-Then continue by creating plots or tables of the outcome(s) of interest and the predictor/exposure/input variables you are most interested in.
-If your dataset is small, you can do that for all variables.
-
-Plots produced here can be scatterplots, boxplots, violinplots, etc. Tables can be simple 2x2 tables or larger ones.
 
@@
 col_names = names(myhrdata)
-num_columns <- 3
+num_columns = 3
 
-num_rows <- ceiling(length(col_names) / num_columns)
+num_rows = ceiling(length(col_names) / num_columns)
 
-summary_matrix <- matrix(col_names, nrow = num_rows, byrow = TRUE)
+summary_matrix = matrix(col_names, nrow = num_rows, byrow = TRUE)
 
-summary_df <- as.data.frame(summary_matrix)
+summary_df = as.data.frame(summary_matrix)
 
-# Print the summary data frame
 print(summary_df)
@@
 
-Gender and Tenure Boxplot
+Tenure Distribution
 
@@
 
-# Data exploration through tables
+```{r}
+p2 = myhrdata %>%
+  ggplot(aes(x = TenureCategory)) +
+  geom_bar(fill = "gray", color = "black") +
+  labs(x = "Tenure", y = "Count", title = "Distribution of Tenure Categories") + 
+  theme_minimal() +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text.x = element_text(angle = 45, hjust = 1)) +
+  guides(fill = FALSE)
 
-Showing a bit of code to produce and save a summary table.
+print(p2)
 
-```{r}
-summary_df = skimr::skim(mydata)
-print(summary_df)
-# save to file
-summarytable_file = here("results", "tables", "summarytable.rds")
-saveRDS(summary_df, file = summarytable_file)
-```
-
-We are saving the results to the `results` folder.
-Depending on how many tables/figures you have, it might make sense to have separate folders for each.
-And/or you could have separate folders for exploratory tables/figures and for final tables/figures.
-Just choose a setup that makes sense for your project and works for you, and provide enough documentation that someone can understand what you are doing.
-
-# Data exploration through figures
-
-Histogram plots for the continuous outcomes.
-
-Height first.
-
-```{r}
-p1 <- mydata %>% ggplot(aes(x=Height)) + geom_histogram() 
-plot(p1)
-figure_file = here("results", "figures","height-distribution.png")
-ggsave(filename = figure_file, plot=p1) 
-```
-
-Now weights.
-
-```{r}
-p2 <- mydata %>% ggplot(aes(x=Weight)) + geom_histogram() 
-plot(p2)
-figure_file = here("results", "figures","weight-distribution.png")
+figure_file = here("results", "figures","tenure-category.png")
 ggsave(filename = figure_file, plot=p2)
@@
 
-Now height as function of weight.
+```{r}
+education_levels <- c("Below College", "College", "Bachelor's", "Master's", "Doctorate")
 
-```{r}
-p3 <- mydata %>% ggplot(aes(x=Height, y=Weight)) + geom_point() + geom_smooth(method='lm')
-plot(p3)
-figure_file = here("results", "figures", "height-weight.png")
+# Convert Education to factor with defined levels
+myhrdata$Education <- factor(myhrdata$Education, levels = education_levels)
+
+# Boxplot of MonthlyIncome by Education
+p3 <- ggplot(myhrdata, aes(x = Education, y = MonthlyIncome)) +
+  geom_boxplot(fill = "gray", color = "black") +
+  labs(x = "Education", y = "Monthly Income", title = "Boxplot of Monthly Income by Education") + 
+  theme_minimal() +
+  theme(panel.background = element_rect(fill = "white"),
+        axis.text.x = element_text(angle = 45, hjust = 1))
+
+# Print the plot
+print(p3)
+
+figure_file = here("results", "figures","income-education.png")
 ggsave(filename = figure_file, plot=p3)
 ```
-
-Once more height as function of weight, stratified by gender. Note that there is so little data, it's a bit silly. But we'll plot it anyway.
-
-```{r}
-p4 <- mydata %>% ggplot(aes(x=Height, y=Weight, color = Gender)) + geom_point() + geom_smooth(method='lm')
-plot(p4)
-figure_file = here("results", "figures", "height-weight-stratified.png")
-ggsave(filename = figure_file, plot=p4) 
-```
-
-# Notes
-
-For your own explorations, tables and figures can be "quick and dirty".
-As long as you can see what's going on, there is no need to polish them.
-That's in contrast to figures you'll produce for your final products (paper, report, presentation, website, etc.).
-Those should look as nice, polished and easy to understand as possible.

### 2024-06-15 code/processing-code/HRprocessingfile.qmd
Project part 2 - second update

@@
 output: html_document
----
-
-# Processing script
-
-This contains the same code and comments/information as `processingcode.R`.
-
-This just shows it as Quarto file, to give you an idea how to do it in a setup that combines code and text in a single file.
-
-See the other Quarto file for my currently preferred approach of pulling code from the R script into the Quarto file.
 
@@
 ```
+```{r}
+tenure_breaks <- seq(0, 40, by = 5)
+tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")
+
+d3 = d2 %>%
+  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))
+
+head(d3)
+```
 
 ```{r}
-HRprocesseddata = d2
+HRprocesseddata = d3
 ```
@@
 ```
-
-# Notes
-
-Removing anyone who had "faulty" or missing data is one approach.
-It's often not the best.
-based on your question and your analysis approach, you might want to do cleaning differently (e.g.
-keep individuals with some missing information).
-
-# Cleaning DELETE WHEN DONE
-
-By inspecting the data as done above, we find some problems that need addressing:
-
-First, there is an entry for height which says "sixty" instead of a number. 
-Does that mean it should be a numeric 60? It somehow doesn't make sense since the weight is 60kg, which can't happen for a 60cm person (a baby).
-Since we don't know how to fix this, we might decide to remove the person.
-This "sixty" entry also turned all Height entries into characters instead of numeric.
-That conversion to character also means that our summary function isn't very meaningful.
-So let's fix that first.
-
-```{r}
-d1 <- rawdata %>% dplyr::filter( Height != "sixty" ) %>% 
-                  dplyr::mutate(Height = as.numeric(Height))
-skimr::skim(d1)
-hist(d1$Height)
-```
-
-Now we see that there is one person with a height of 6.
-That could be a typo, or someone mistakenly entered their height in feet.
-Since we unfortunately don't know, we might need to remove this person, which we'll do here.
-
-```{r}
-d2 <- d1 %>% dplyr::mutate( Height = replace(Height, Height=="6",round(6*30.48,0)) )
-skimr::skim(d2)
-```
-
-Height values seem ok now.
-
-Now let's look at the `Weight` variable. There is a person with weight of 7000, which is impossible, and one person with missing weight.
-To be able to analyze the data, we'll remove those individuals as well.
-
-```{r}
-d3 <- d2 %>%  dplyr::filter(Weight != 7000) %>% tidyr::drop_na()
-skimr::skim(d3)
-```
-
-Now checking the `Gender` variable. Gender should be a categorical/factor variable but is loaded as character. We can fix that with simple base R code to mix things up.
-
-```{r}
-d3$Gender <- as.factor(d3$Gender)  
-skimr::skim(d3)
-```
-
-Now we see that there is another NA, but it's not `NA` from R, instead it was loaded as character and is now considered as a category.
-Well proceed here by removing that individual with that NA entry. Since this keeps an empty category for Gender, I'm also using droplevels() to get rid of it.
-
-```{r}
-d4 <- d3 %>% dplyr::filter( !(Gender %in% c("NA","N")) ) %>% droplevels()
-skimr::skim(d4)
-```
-
-All done, data is clean now.
-
-Let's assign at the end to some final variable, this makes it easier to add further cleaning steps above.

### 2024-06-15 products/manuscript/Manuscript.qmd
Project part 2 - second update

@@
 
-**Removing Column**
+**Removing Columns**
 
@@
 
-**Adding Labels**
+**Adding Descriptive Labels**
 
@@
 head(d2)
+```
+
+**Tenure Categorization**
+
+To enhance the visualization of tenure distribution, I opted to introduce a new variable categorizing employee tenure.
+This approach allows us to identify the predominant tenure group, offering insights into potential turnover risks.
+
+```{r, results='hide'}
+tenure_breaks <- seq(0, 40, by = 5)
+tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")
+
+d3 = d2 %>%
+  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))
 ```
@@
 
-### **Gender**
-
-The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%.
-
-**Distribution**
-
-```{r}
-#| label: col-hrgender
-#| tbl-cap: "Employee Gender"
-#| echo: FALSE
-gender_table=readRDS("../../results/tables/HRgendertable.rds")
-knitr::kable(gender_table)
-```
-
 ### **Employee Tenure (in years)**
@@
 
-**MAYBE ADD LABEL FOR TENURE?**
+**Tenure category**
 
-### **Education** MOVE TO EDA
+More than half of the employees have tenure of less than 5 years, suggesting a relatively young workforce.
+Shorter tenure could potentially indicate a higher turnover risk, although a comprehensive analysis incorporating other variables is necessary to fully understand the situation.
 
-During data cleaning, we'll use the data dictionary to decode each field.
-For instance, the education field is labeled from 1 to 5, representing different educational levels: 1 = "Below College," 2 = "College," 3 = "Bachelor's," 4 = "Master's," and 5 = "Doctorate." Notably, about 40% of the dataset holds a Bachelor's degree.
+```{r}
+#| label: fig-tenureCategory
+#| fig-cap: "Tenure Category"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","tenure-category.png"))
+```
+
+### **Gender**
+
+The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%.
+
+**Distribution**
+
+```{r}
+#| label: tbl-hrgender
+#| tbl-cap: "Employee Gender"
+#| echo: FALSE
+gender_table=readRDS("../../results/tables/HRgendertable.rds")
+knitr::kable(gender_table)
+```
+
+### **Monthly Income and Education**
+
+I'm interested in examining which educational group experiences higher attrition rates.
+It's evident from the data that median salaries increase with higher levels of education, but there are also numerous outliers present.
+This visualization serves to highlight the relationship between monthly income and educational background, which could provide insights into how these factors influence employee turnover.
 
 ```{r }
-
+#| label: fig-Income-Education
+#| fig-cap: "Monthly Income by Education Level"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","income-education.png"))
 ```

### 2024-06-19 code/eda-code/HReda.qmd
Project part 3 - First Update

@@
 
-# HR Data Exploration Through Tables
+## HR Data Exploration Through Tables
 
@@
 
+## Data Summary
+
+Basic summary of data.
 ```{r}
@@
 
-Gender Table
+## Gender Table
 
 ```{r}
-#summary_df = skimr::skim(myhrdata)
-#print(summary_df)
 
@@
   mutate(`Total Count` = n,
-         `% of Total` = (n / sum(n)) * 100) %>%
-  select(Gender, `Total Count`, `% of Total`)
+         `% of Total` = (n / sum(n)) * 100) %>% #creating new column for percentage of total
+  select(Gender, `Total Count`, `% of Total`) #selecting only 3 variables
 
@@
 
-Tenure Distribution
+## Tenure Distribution
 
+Visualizing tenure to find distribution.
 ```{r}
@@
   labs(x = "Tenure", y = "Count") + 
-  theme_minimal() +
-  theme(panel.background = element_rect(fill = "white"))
+  theme_minimal()
 
@@
 
+## Distribution of Tenure
+
 ```{r}
@@
   labs(x = "Tenure", y = "Count", title = "Distribution of Tenure Categories") + 
-  theme_minimal() +
-  theme(panel.background = element_rect(fill = "white"),
-        axis.text.x = element_text(angle = 45, hjust = 1)) +
-  guides(fill = FALSE)
+  theme_minimal()
 
@@
 
+## Education vs Monthly Income
+
 ```{r}
-education_levels <- c("Below College", "College", "Bachelor's", "Master's", "Doctorate")
-
-# Convert Education to factor with defined levels
-myhrdata$Education <- factor(myhrdata$Education, levels = education_levels)
-
-# Boxplot of MonthlyIncome by Education
-p3 <- ggplot(myhrdata, aes(x = Education, y = MonthlyIncome)) +
+p3 = ggplot(myhrdata, aes(x = Education, y = MonthlyIncome)) +
   geom_boxplot(fill = "gray", color = "black") +
   labs(x = "Education", y = "Monthly Income", title = "Boxplot of Monthly Income by Education") + 
-  theme_minimal() +
-  theme(panel.background = element_rect(fill = "white"),
-        axis.text.x = element_text(angle = 45, hjust = 1))
+  theme_minimal()
 
-# Print the plot
 print(p3)
@@
 ```
+
+## Age vs Attrition
+
+```{r}
+p5 = ggplot(myhrdata, aes(x = Age, fill = Attrition)) +
+  geom_histogram(binwidth = 1, position = "dodge") +
+  scale_fill_manual(values = c("No" = "grey", "Yes" = "#595959")) + #selecting bar colors
+  labs(title = "Age vs Attrition", x = "Age", y = "Count") +
+  theme_minimal()
+
+print(p5)
+
+figure_file = here("results", "figures","age-attrition.png")
+ggsave(filename = figure_file, plot=p5)
+```
+
+```{r}
+p4 = ggplot(myhrdata, aes(x = Department, fill = Attrition)) +
+  geom_bar(position = "dodge", color = "black") +
+  scale_fill_manual(values = c("No" = "grey", "Yes" = "#595959")) + #selecting bar colors
+  labs(title = "Department vs Attrition", x = "Department", y = "Count") +
+  theme_minimal()
+
+print(p4)
+
+figure_file = here("results", "figures","department-attrition.png")
+ggsave(filename = figure_file, plot=p4)
+```

### 2024-06-19 code/processing-code/HRprocessingfile.qmd
Project part 3 - First Update

@@
 output: html_document
+---
 
@@
 
-**NEED TO UPDATE THE EDUCATION FIELD**
-
 # Cleaning HR Data
@@
 
+Saving updated dataset into HRprocesseddata
 ```{r}

### 2024-06-19 products/manuscript/Manuscript.qmd
Project part 3 - First Update

@@
 
-#hr_data = read.csv("C:/Users/crew_/OneDrive/Escritorio/Guille/UTSA Masters/MSDA/DA6833/PracticumII-Course/guillermogallardo-final-project/data/raw-data/IBM HR Analytics Data.csv", header = TRUE) #not synced with OneDrive but my computer added that folder when I entered my email when I bought the computer
 ```
@@
 
+### **Age vs Attrition**
+
+In the bar chart, it is evident that attrition peaks among employees aged between 28 and 32.
+This trend suggests that individuals in this age group may be more likely to leave the company.
+Possible reasons for increased turnover in this demographic could include career advancement opportunities elsewhere, desire for higher compensation, or life changes such as starting a family.
+
+```{r}
+#| label: fig-Age-Attrition
+#| fig-cap: "Attrition by Employee Age"
+#| echo: FALSE
+knitr::include_graphics(here("results","figures","age-attrition.png"))
+```
+
 ## Correlation Analysis
@@
 I am interested in using the demographic details, job attributes and survey data to uncover pattern (if any) that highlight this issue .
+
+```{r}
+
+```
 

### 2024-06-19 code/analysis-code/statistical-analysis.R

@@ -8,18 +8,52 @@
 library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
+library(tidyverse)
+library(tidymodels)
+
+update.packages(ask = FALSE)
 
 #path to data
 #note the use of the here() package and not absolute paths
-data_location <- here::here("data","processed-data","processeddata.rds")
+data_location = here::here("data","processed-data","HRprocesseddata.rds")
 
 #load data. 
-mydata <- readRDS(data_location)
+mydata = readRDS(data_location)
 
 
 ######################################
 #Data fitting/statistical analysis
 ######################################
+
+#### Linear Model Job Satisfaction ~ Department
+
+#lmfitHR = lm(JobSatisfaction~Age, data = mydata)
+
+#summary(lmfitHR)
+
+glm_spec <- logistic_reg() %>%
+  set_engine("glm")
+
+# Split the data
+set.seed(123)
+data_split <- initial_split(mydata, prop = 0.75, strata = Attrition)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+
+# Fit the model
+glm_fit <- glm_spec %>%
+  fit(Attrition ~ Age + DistanceFromHome + Education + EnvironmentSatisfaction + JobInvolvement + JobLevel + JobRole + StockOptionLevel + TotalWorkingYears + TrainingTimesLastYear + WorkLifeBalance + YearsAtCompany + YearsInCurrentRole + YearsSinceLastPromotion + YearsWithCurrManager + TenureCategory, data = train_data)
+
+# Model summary
+summary(glm_fit$fit)
+
+# Predict and evaluate the model
+glm_preds <- predict(glm_fit, test_data, type = "prob")
+glm_roc <- roc_auc(truth = test_data$Attrition, .pred_Yes = glm_preds$.pred_Yes)
+
+# Save the results
+
+
 
 ############################
 #### First model fit

### 2024-06-22 products/manuscript/Manuscript.qmd
Project part 3 - second update

@@
 
--   What strategies can be implement to enhance employee retention and improve the overall employee experience?
+-   What strategies can we implement to enhance employee retention and improve the overall employee experience?
 
@@
 The dataset for this project was sourced from [Kaggle.com](https://www.kaggle.com/datasets/pavansubhasht/ibm-hr-analytics-attrition-dataset) and was created by IBM Data Scientists as a fictional representation of HR data.
-It consists of 1,470 rows and 35 variables that cover demographic details, job characteristics, compensation, performance metrics, and employee satisfaction data.
+It consists of 1,470 observations and 35 variables that cover demographic details, job characteristics, compensation, performance metrics, and employee satisfaction data.
 This data serves as the foundation for analyzing factors contributing to employee turnover and developing a predictive model.
@@
 In the code below, we use the *library(here)* function to set the file path dynamically, so our script can find and access the data files regardless of the current working directory.
+You can update this code to point to a specific folder path if needed but this library will make it easier to reproduce the file with the existing folders.
 
@@
 ```{r, results='hide'}
-hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx")
-rawdata = readxl::read_excel(hr_data)
+hr_data = here::here("data","raw-data","IBM HR Analytics Data.xlsx") #file location
+rawdata = readxl::read_excel(hr_data) #reading the file
 ```
@@
 d1 = rawdata %>% 
-  select(-DailyRate, -HourlyRate, -MonthlyRate)
+  select(-DailyRate, -HourlyRate, -MonthlyRate) # the (-) will remove these variables from the dataset
 
-head(d1)
+head(d1) #priting the first 6 observations
 ```
@@
 ```{r, results='hide'}
-tenure_breaks <- seq(0, 40, by = 5)
+tenure_breaks <- seq(0, 40, by = 5) #defining breaks for categories
 tenure_labels <- paste0(tenure_breaks[-length(tenure_breaks)], "-", tenure_breaks[-1], " years")
@@
 d3 = d2 %>%
-  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE))
+  mutate(TenureCategory = cut(YearsAtCompany, breaks = tenure_breaks, labels = tenure_labels, include.lowest = TRUE)) #creating a new column
 ```
@@
 ```{r, results='hide'}
-HRprocesseddata = d2
+HRprocesseddata = d2 #creating final object
 
-save_data_location = here::here("data","processed-data","HRprocesseddata.rds")
+save_data_location = here::here("data","processed-data","HRprocesseddata.rds") #using the here library for the location of new file
 saveRDS(HRprocesseddata, file = save_data_location)
@@
 This analysis will help us understand the data available to us and guide our strategy to predict employee attrition within the organization.
+The variables shown are derived from the dataset that we cleaned in the previous step.
+This includes new columns, updated values, and excludes columns that we removed.
 
@@
 
-This dataset contains 35 variables, consisting of both character and integer types, with several character variables categorized as categorical.
+This dataset contains 33 variables, consisting of both character and integer types, with several character variables categorized as categorical.
 It provides valuable insights into employees, including demographic information such as age, gender, and marital status.
@@
 This provide quick insights into the typical duration of employees within the organization, and help us identify patterns that may influence employees to resign.
+We could concentrate on these groups to identify commonalities and understand the reasons behind their departure from the organization.
 
@@
 The dataset consists of 1470 employees, with a gender distribution showing that males constitute the majority at 60%, while females make up 40%.
+This also highlights the gender disparity, and it would be interesting to explore this gap by department and examine attrition rates between the two groups.
 
@@
 
-## Correlation Analysis
+Some of these variables may explain why employees are leaving.
+Understanding the employee profile by examining age, gender, department, and job satisfaction can reveal the reasons behind their departure.
+By creating profiles based on these insights, we can develop targeted retention strategies to address the specific needs and concerns of these employees.
 
-As we explore various analysis methods in our class, my focus is on finding potential correlation between variables and employee attrition.
-I am interested in using the demographic details, job attributes and survey data to uncover pattern (if any) that highlight this issue .
-
-```{r}
-
-```
-
-## Predictive Modeling
-
-My main goal is to predict when an employee might leave the company.
-I am interested in using predictive modeling techniques (also part of my other summer class) to see if we can predict attrition.
-Once I've identified potential departures, we can use this findings to create strategies to prevent or reduce attrition.
+## Additional Analysis

### 2024-06-22 code/analysis-code/statistical-analysis.R

@@ -29,29 +29,6 @@ mydata = readRDS(data_location)
 
 #lmfitHR = lm(JobSatisfaction~Age, data = mydata)
 
-#summary(lmfitHR)
-
-glm_spec <- logistic_reg() %>%
-  set_engine("glm")
-
-# Split the data
-set.seed(123)
-data_split <- initial_split(mydata, prop = 0.75, strata = Attrition)
-train_data <- training(data_split)
-test_data <- testing(data_split)
-
-# Fit the model
-glm_fit <- glm_spec %>%
-  fit(Attrition ~ Age + DistanceFromHome + Education + EnvironmentSatisfaction + JobInvolvement + JobLevel + JobRole + StockOptionLevel + TotalWorkingYears + TrainingTimesLastYear + WorkLifeBalance + YearsAtCompany + YearsInCurrentRole + YearsSinceLastPromotion + YearsWithCurrManager + TenureCategory, data = train_data)
-
-# Model summary
-summary(glm_fit$fit)
-
-# Predict and evaluate the model
-glm_preds <- predict(glm_fit, test_data, type = "prob")
-glm_roc <- roc_auc(truth = test_data$Attrition, .pred_Yes = glm_preds$.pred_Yes)
-
-# Save the results
 
 
 
@@ -87,4 +64,3 @@ print(lmtable2)
 table_file2 = here("results", "tables", "resulttable2.rds")
 saveRDS(lmtable2, file = table_file2)
 
-  

### 2024-07-11 products/manuscript/Manuscript.qmd
Project part 3 - statistical analysis

@@
 saveRDS(HRprocesseddata, file = save_data_location)
+
 ```
@@
 
-### **Employee Tenure (in years)**
+### Employee Tenure (in years)
 
@@
 
-**Tenure category**
+### Tenure category
 
@@
 
-### **Gender**
+### Gender
 
@@
 
-### **Monthly Income and Education**
+### Monthly Income and Education
 
@@
 
-### **Age vs Attrition**
+### Age vs Attrition
 
@@
 
-## Additional Analysis
+## Statistical Analysis Analysis
+
+### Logistic Regression with Age
+
+I conducted logistic regressions to explore predictors of Attrition among employees.
+Age emerged as a significant factor, with each additional year decreasing Attrition odds by approximately 0.052 (p < 0.001), underscoring its impact on turnover.
+
+These findings emphasize the need for age-sensitive strategies, such as mentorship programs for knowledge transfer, flexible work arrangements, and tailored career development opportunities.
+Leveraging experience can foster commitment and reduce turnover among senior employees.
+However, limitations include the model's inability to capture all turnover influencers, like career aspirations or economic factors.
+
+```{r}
+#| label: tbl-logAge
+#| tbl-cap: "Logistic Regression - Age"
+#| echo: FALSE
+logAge_table=readRDS("../../results/tables/logistic_age.rds")
+knitr::kable(logAge_table)
+
+```
+
+### Logistic Regression with Income
+
+Money plays a crucial role in life decisions, including work-related choices.
+I examined Monthly Income to understand its impact on Attrition among employees.
+The results show that higher Monthly Income is associated with lower odds of Attrition.
+Specifically, for each unit increase in Monthly Income, the odds of Attrition decrease by about 0.0001271 (p < 0.001), highlighting its significant influence on turnover.
+
+To leverage these findings, organizations can consider implementing competitive salary structures, performance-based incentives, and financial wellness programs.
+These initiatives can enhance employee satisfaction and retention, particularly among those sensitive to income-related factors.
+```{r}
+#| label: tbl-logInc
+#| tbl-cap: "Logistic Regression - Income"
+#| echo: FALSE
+logInc_table=readRDS("../../results/tables/logistic_income.rds")
+knitr::kable(logInc_table)
+
+```
+
+### Logistic Regresion with Age, Income, and Job Satisfaction
+
+I used a logistic regression model to study how Age, Monthly Income, and Job Satisfaction collectively affect Attrition among employees.
+This approach allowed me to explore their combined impact on turnover.
+The analysis revealed that older employees, those with higher incomes, and greater job satisfaction are significantly less likely to leave.
+Specifically, each additional year of Age reduces Attrition likelihood by about 0.0328 (p < 0.001), with similar significant impacts observed for Monthly Income and Job Satisfaction.
+```{r}
+#| label: tbl-glm
+#| tbl-cap: "GLM - Age, Income, and Job Satisfaction"
+#| echo: FALSE
+glm_table=readRDS("../../results/tables/glm_model.rds")
+knitr::kable(glm_table)
+```
+
+## Next Steps
+
+We must investigate more predictors of attrition to identify root causes and develop essential strategies for retaining employees and improving the overall employee experience.

### 2024-07-11 code/analysis-code/statistical-analysis.R

@@ -10,57 +10,152 @@ library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
 library(tidyverse)
 library(tidymodels)
+library(caret) # For modeling and evaluation
+library(e1071)  
+library(rsample)
+library(parsnip)
 
-update.packages(ask = FALSE)
-
-#path to data
-#note the use of the here() package and not absolute paths
+####################################
+# Loading Data
 data_location = here::here("data","processed-data","HRprocesseddata.rds")
+hr_data = readRDS(data_location)
 
-#load data. 
-mydata = readRDS(data_location)
-
-
-######################################
-#Data fitting/statistical analysis
-######################################
-
-#### Linear Model Job Satisfaction ~ Department
-
-#lmfitHR = lm(JobSatisfaction~Age, data = mydata)
+####################################
 
 
+#Bivariate
 
+#####################################
+## Age Logistic Regression
+age_glm = glm(Attrition ~ Age, data = hr_data, family = binomial())
+summary(age_glm)
 
-############################
-#### First model fit
-# fit linear model using height as outcome, weight as predictor
-
-lmfit1 <- lm(Height ~ Weight, mydata)  
-
-# place results from fit into a data frame with the tidy function
-lmtable1 <- broom::tidy(lmfit1)
-
+## Age Export table
+###tidy function
+lgAge = broom::tidy(age_glm)
+lgAge$p.value <- format(lgAge$p.value, digits = 6)
 #look at fit results
-print(lmtable1)
-
+print(lgAge)
 # save fit results table  
-table_file1 = here("results", "tables", "resulttable1.rds")
-saveRDS(lmtable1, file = table_file1)
+table_lgAge = here("results", "tables", "logistic_age.rds")
+saveRDS(lgAge, file = table_lgAge)
+####################################
 
-############################
-#### Second model fit
-# fit linear model using height as outcome, weight and gender as predictor
 
-lmfit2 <- lm(Height ~ Weight + Gender, mydata)  
+####################################
+## Income Logistic Regression
+inc_glm = glm(Attrition ~ MonthlyIncome, data = hr_data, family = binomial())
+summary(inc_glm)
 
-# place results from fit into a data frame with the tidy function
-lmtable2 <- broom::tidy(lmfit2)
-
+## Income Export table
+###tidy function
[…]

### 2024-07-13 products/manuscript/Manuscript.qmd
Project part 4 - Statistical Analysis

@@
 
-The primary objective of this project is to identify the key factors contributing to employee turnover and to develop a model that can predict potential departures within a company.
-By anticipating these departures, we can create strategies for employee retention and enhance the overall employee experience.
+The primary objective of this project is to identify the key factors contributing to employee turnover by developing a model that can provide the predictors that influence employees to leave the organization.
+By identifying this factors, we can create strategies for employee retention and enhance the overall employee experience.
 
@@
 
-**Loading Data for Cleaning Process**
+## Loading Data for Cleaning Process
 
@@
 
-**Removing Columns**
+## Removing Columns
 
@@
 
-**Adding Descriptive Labels**
+## Adding Descriptive Labels
 
@@
 
-**Tenure Categorization**
+## Tenure Categorization
 
@@
 
-**Saving File**
+## Saving File
 
@@
 I conducted logistic regressions to explore predictors of Attrition among employees.
-Age emerged as a significant factor, with each additional year decreasing Attrition odds by approximately 0.052 (p < 0.001), underscoring its impact on turnover.
+Age emerged as a significant factor, with each additional year decreasing Attrition odds by approximately 0.052 (p \< 0.001), underscoring its impact on turnover.
 
@@
 The results show that higher Monthly Income is associated with lower odds of Attrition.
-Specifically, for each unit increase in Monthly Income, the odds of Attrition decrease by about 0.0001271 (p < 0.001), highlighting its significant influence on turnover.
+Specifically, for each unit increase in Monthly Income, the odds of Attrition decrease by about 0.0001271 (p \< 0.001), highlighting its significant influence on turnover.
 
@@
 These initiatives can enhance employee satisfaction and retention, particularly among those sensitive to income-related factors.
+
 ```{r}
@@
 The analysis revealed that older employees, those with higher incomes, and greater job satisfaction are significantly less likely to leave.
-Specifically, each additional year of Age reduces Attrition likelihood by about 0.0328 (p < 0.001), with similar significant impacts observed for Monthly Income and Job Satisfaction.
+Specifically, each additional year of Age reduces Attrition likelihood by about 0.0328 (p \< 0.001), with similar significant impacts observed for Monthly Income and Job Satisfaction.
+
 ```{r}
@@
 
-## Next Steps
+### Random Forest Model
 
-We must investigate more predictors of attrition to identify root causes and develop essential strategies for retaining employees and improving the overall employee experience.
+For our last model, I choose a Random Forest due to its capability to manage intricate, nonlinear relationships among numerous variables without overfitting.
+This model excels in accuracy and offers valuable insights into predictor importance, helping us identify the most influential factors affecting attrition.
+
+For this model, I did the following:
+
+-   Split the dataset into 80% training and 20% testing to evaluate model performance. This suggestion was taken from my predictive modeling class.
+
+-   We then balanced the training data to address class imbalanced which can affect the model accuracy.
+There were many more examples of non attrition than attrition so we needed to address them for the model.
+
+The call below specifies the model's formula, data, and parameters used, while noting it's a classification Random Forest predicting Attrition **(Yes 1 /No 0)** outcomes with 1000 trees.
+Each tree uses 5 variables per split to enhance stability and accuracy, balancing computational resources against model variance.
+
+```{r}
+#| label: tbl-rfmodel
+#| tbl-cap: "Random Forest Model"
+#| echo: FALSE
+rf_model=readRDS("../../results/tables/rf_summary.rds")
+knitr::kable(rf_model)
+```
+
+#### Random Forest Model Results
+
+Performance: The Random Forest model demonstrated excellent predictive capabilities, with an OOB error rate of just 3.95%, indicating high accuracy.
+The confusion matrix confirmed strong performance, particularly in identifying true positives and negatives, with low false positive and false negative rates.
+
+**Class Error:**
+
+-   0.0141844 for '0' (No Attrition): Very low error rate, indicating high accuracy for predicting non-attrition.
+
+    -   73 True Negatives (TN): Non-attrition cases correctly identified.
+
+    -   14 False Positives (FP): Non-attrition cases incorrectly predicted as attrition.
+
+-   0.1052632 for '1' (Attrition): Higher error rate, suggesting room for improvement in predicting actual attrition cases.
+
+    -   40 False Negatives (FN): Attrition cases missed by the model.
+
+    -   340 True Positives (TP): Attrition cases correctly identified.
+
+### Random Forest Top 5 Predictors
+
+**Top 5 Predictors:**
+
+1.
+Overtime: has the highest decrease in accuracy and Gini when excluded, indicating it's a crucial predictor of attrition.
+High values suggest that employees working overtime are significantly more likely to leave.
+
+2.
+Monthly Income: This variable strongly influences attrition, implying that employees with certain income levels may be more prone to leaving, possibly due to pay dissatisfaction or better opportunities elsewhere.
+
+3.  Age: impacts attrition decisions, with potential reasons including generational expectations, life stage career goals, and job fit.
+
+4.  Job Satisfaction: A critical predictor indicating that lower job satisfaction is strongly associated with higher attrition.
+
+5.  Distance from Home: Longer commutes are linked to higher attrition, suggesting that travel time affects employee satisfaction and retention.
+
+```{r}
+#| label: tbl-rfpred
+#| tbl-cap: "Random Forest Top Predictors"
+#| echo: FALSE
+rf_5pred=readRDS("../../results/tables/rd_top_predictors.rds")
+knitr::kable(rf_5pred)
+```
+
+# Strategies to Prevent Attrition
+
+Based on insights from the models, it's evident that factors such as overtime, monthly income, age, job satisfaction, and distance from home significantly influence employees' decisions to leave the organization.
+Armed with this information, we can create strategies focus at reducing employee turnover.
+
+Proposed strategies include:
+
+-   OverTime Management: Implement strict overtime policies and ensure workload distribution is balanced. Encouraging time-off and providing support during peak times can prevent burnout.
+
+-   Compensation and Benefits: Regular market analysis to ensure competitive pay scales and comprehensive benefits tailored to meet the diverse needs of the workforce can help in retaining talent, particularly those influenced by financial incentives.
+
+-   Age-Adaptive HR Policies: Designing age-specific policies that cater to the unique needs and preferences of different generational cohorts within the workplace.
+For example, younger employees might value professional development and upskilling opportunities more than their older counterparts, who might prioritize job security and healthcare benefits.
+
+-   Enhancing Job Satisfaction: Focus on creating a positive work environment where employees feel valued and have clear career paths.
+This can include team-building activities, transparent communication from management, and continuous professional development.
+
+-   Flexible Working Arrangements: For employees burdened by long commutes, flexible working arrangements can be a game-changer.
+This could range from offering telecommuting options to establishing satellite offices closer to where clusters of employees live.
+
+# Conclusion
+
+Building on these strategies, it's clear that addressing key factors like overtime, monthly income, age, job satisfaction, and distance from home can mitigate attrition risks.
+However, retaining talent involves ongoing efforts beyond initial strategies.
+Establishing a robust feedback system for continuous improvement, fostering a culture of open communication and trust, and regularly reviewing and adapting policies based on employee feedback can further strengthen retention efforts.
+Investing in leadership development programs to empower managers with skills to support and engage their teams effectively can also play a crucial role.
+By integrating these approaches with a commitment to fostering a supportive and inclusive workplace culture, organizations can create an environment where employees are motivated to stay and thrive, ultimately reducing turnover and fostering long-term organizational success.

### 2024-07-13 code/analysis-code/statistical-analysis.R

@@ -5,7 +5,6 @@
 #and saves the results to the results folder
 
 #load needed packages. make sure they are installed.
-library(ggplot2) #for plotting
 library(broom) #for cleaning up output from lm()
 library(here) #for data loading/saving
 library(tidyverse)
@@ -16,16 +15,21 @@ library(rsample)
 library(parsnip)
 
 ####################################
+####################################
 # Loading Data
 data_location = here::here("data","processed-data","HRprocesseddata.rds")
 hr_data = readRDS(data_location)
-
+hr_data$Attrition = ifelse(hr_data$Attrition == "Yes", 1, 0)
+table(hr_data$Attrition)
+####################################
 ####################################
 
 
-#Bivariate
 
-#####################################
+
+#Bivariate
+####################################
+####################################
 ## Age Logistic Regression
 age_glm = glm(Attrition ~ Age, data = hr_data, family = binomial())
 summary(age_glm)
@@ -33,15 +37,19 @@ summary(age_glm)
 ## Age Export table
 ###tidy function
 lgAge = broom::tidy(age_glm)
-lgAge$p.value <- format(lgAge$p.value, digits = 6)
+lgAge$p.value = format(lgAge$p.value, digits = 6)
 #look at fit results
 print(lgAge)
 # save fit results table  
 table_lgAge = here("results", "tables", "logistic_age.rds")
 saveRDS(lgAge, file = table_lgAge)
 ####################################
+####################################
 
 
+
+
+####################################
 ####################################
 ## Income Logistic Regression
 inc_glm = glm(Attrition ~ MonthlyIncome, data = hr_data, family = binomial())
@@ -50,29 +58,35 @@ summary(inc_glm)
 ## Income Export table
 ###tidy function
 lgInc = broom::tidy(inc_glm)
-lgInc$p.value <- format(lgInc$p.value, digits = 6)
+lgInc$p.value = format(lgInc$p.value, digits = 6)
 #look at fit results
 print(lgInc)
 # save fit results table  
 table_lgInc = here("results", "tables", "logistic_income.rds")
 saveRDS(lgInc, file = table_lgInc)
 ####################################
+####################################
+
 
 
 ####################################
+####################################
 ## Chi-squared
-education_table <- table(hr_data$Attrition, hr_data$EducationField)
-chisq_test <- chisq.test(education_table)
[…]

### 2024-07-30 products/manuscript/Manuscript.qmd
final project

@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/apa.csl
 ---

### 2024-07-30 code/analysis-code/statistical-analysis.R

@@ -1,5 +1,5 @@
 ###############################
-# analysis script
+# Analysis script
 #
 #this script loads the processed, cleaned data, does a simple analysis
 #and saves the results to the results folder
